Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-03-29 WSSSPE1_report/WSSSPE1_summary_report.tex
Very minor edits to Policy section.
Mostly typos and citations.

One group of papers presented frameworks that were primarily about characteristics of software artifacts, connecting with the long discourse on software quality. This approach is realized in adjectives that can be applied to pieces of software but might also extend to describe software projects.  Thus Calero and colleagues propose adding elements to the ISO standards for measuring software quality: proposing an additional dimension of quality they call perduability or ability to ``last over time'' made up of three characteristics: reliability, maintainability and adaptability. This shows some overlap with the framework employed by Venters et al.~\cite{Venters_WSSSPE} who employ the adjectives of ``extensibility, interoperability, maintainability, portability, reusability and scalability,'' anticipating the sorts of work that people would need to do with a software artifact in the future, ``as stakeholders requirements, technology and environments evolve and change.'' Venters et al. argue that these choices are really made very early because they relate to the architecture of the software and involve trade offs that ought to be analyzed alongside other attributes of the software artifact. Lenhardt et al compare the software lifecycle to the data lifecycle to argue for the inclusion of metadata throughout a piece of software's life (discussing, for example, how it has been built and tested and what ``data in'' and ``data out'' has been considered.) In addition their analogy suggests that the software lifecycle might add a phase of ``preservation'' and draw on understandings of what that involves from studies of data.  In sum, then, these frameworks focused on what needs to be accomplished to have more sustainable software.

A second theme in these papers was the continued availability of resources to accomplish the goals of sustainability. The elements of these frameworks focused far more on the organization of a software project than they did on characteristics of the artifact itself (although it is certainly true that the adjectives discussed above could be applied to a software project). For example Pierce et al focus on the way that the project is run, particularly in terms of how those involved communicate and jointly set priorities, a process they call governance. In particular they argue that because sustainability relates to having ongoing resources, governance must be open to receive diverse input (by occurring online, asynchronously, and publicly) and thus have the potential to ``transform passive users into active contributors''. They argue that the Apache Software Foundation's incubation process teaches this and could be learned from by projects throughout scientific software.  Katz and Proctor also discuss governance, describing two modes: ``top-down'' and ``bottom-up'' governance~\cite{Katz_WSSSPE}.  They place governance alongside questions of the artifact (``technical''), and questions of the origins of resources, in terms of who is funding the work surrounding a piece of software (``the political'') and the manner in which resources come to the project initially (commercial, open source, closed partnerships, grant funded) and long-term (all four plus paid support).

@@

Another approach is to try to reduce the difficulty of citing software for authors, acknowledging that authors are often working with software that itself wraps other software and therefore hides that software. Knepley et. al. approach this by proposing a mechanism by which the software itself, after it has run, provides the user with a set of citations that are relevant to the code actually run. They describe a prototype implementation whereby the citations are embedded in libraries and reported along with the results, via a commandline interface~\cite{Knepley_WSSSPE}. Discussion highlighted the difficulty that attempting to acknowledge the contributions of all pieces of dependent code within a paper faces the difficulty of creating very long citation lists, straining the analogy of code used to papers cited. Katz approaches this issue by proposing a system of ``transitive credit,'' recording dependencies and relative contributions outside particular papers, relieving authors from the responsibility of acknowledging each and every dependency. Instead authors would acknowledge the percentage contribution of the software they used directly and an external system would then be able to recursively allocate that credit to those who had provided dependencies. Finally Priem and Piwowar argued that machine learning techniques could examine the body of published literature and extract mentions of software, coping with the multitude of informal ways in which authors mention software they have used~\cite{Priem_WSSSPE}.  Discussion included turning the emphasis from asking users to improve their citation of software contributions, to ask how projects producing software might monitor the literature to improve their ability to show impact. Michael McLennan described the approach taken by the NanoHub project to scan the literature using keywords and names of known users to discover papers that are likely to have used their software and platform, then to have graduate students read each paper, highlighting any mention and perhaps following up with the authors to locate stories of impact.  Work since the workshop has described this process at \textit{publications.wikia.com}.

Acknowledgement in science does not come only in publications, of course. A key location for visibility might be in the grant funding process, both in bio-sketches and in grant project reporting. Some progress has been made here. Representatives of the NSF at the meeting emphasized these opportunities and encouraged participants to take advantage of them. Nonetheless, there was skepticism that peer review panels would value these contributions in the same ways as publications).

Priem and Piwowar argued for additionally looking beyond publications and drawing on evidence of contribution, and impact, recorded in other online systems, such as source code repositories like github, both code contributions, downloads of code and dependencies as well as conversations about software (mailing lists, twitter and beyond)~\cite{Priem_WSSSPE}. They argue for providing scholars with flexible resources so that they can tell their own stories in the manner most appropriate for them and their audiences, a principle of the ``alt.metrics'' approach.

>>>>

One group of papers presented frameworks that were primarily about characteristics of software artifacts, connecting with the long discourse on software quality. This approach is realized in adjectives that can be applied to pieces of software but might also extend to describe software projects.  Thus Calero and colleagues~\cite{Calero_WSSSPE} propose adding elements to the ISO standards for measuring software quality: proposing an additional dimension of quality they call perdurability or ability to ``last over time'' made up of three characteristics: reliability, maintainability and adaptability. This shows some overlap with the framework employed by Venters et al.~\cite{Venters_WSSSPE} who employ the adjectives of ``extensibility, interoperability, maintainability, portability, reusability and scalability,'' anticipating the sorts of work that people would need to do with a software artifact in the future, ``as stakeholders requirements, technology and environments evolve and change.'' Venters et al. argue that these choices are really made very early because they relate to the architecture of the software and involve trade offs that ought to be analyzed alongside other attributes of the software artifact. Lenhardt et al~\cite{Lenhardt_WSSSPE} compare the software lifecycle to the data lifecycle to argue for the inclusion of metadata throughout a piece of software's life (discussing, for example, how it has been built and tested and what ``data in'' and ``data out'' has been considered). In addition, their analogy suggests that the software lifecycle might add a phase of ``preservation'' and draw on understandings of what that involves from studies of data.  In sum, then, these frameworks focused on what needs to be accomplished to have more sustainable software.

A second theme in these papers was the continued availability of resources to accomplish the goals of sustainability. The elements of these frameworks focused far more on the organization of a software project than they did on characteristics of the artifact itself (although it is certainly true that the adjectives discussed above could be applied to a software project). For example Pierce et al~\cite{Pierce_WSSSPE} focus on the way that the project is run, particularly in terms of how those involved communicate and jointly set priorities, a process they call governance. In particular they argue that because sustainability relates to having ongoing resources, governance must be open to receive diverse input (by occurring online, asynchronously, and publicly) and thus have the potential to ``transform passive users into active contributors''. They argue that the Apache Software Foundation's incubation process teaches this and could be learned from by projects throughout scientific software.  Katz and Proctor~\cite{Katz_WSSSPE} also discuss governance, describing two modes: ``top-down'' and ``bottom-up'' governance.  They place governance alongside questions of the artifact (``technical''), and questions of the origins of resources, in terms of who is funding the work surrounding a piece of software (``the political'') and the manner in which resources come to the project initially (commercial, open source, closed partnerships, grant funded) and long-term (all four plus paid support).

@@

Another approach is to try to reduce the difficulty of citing software for authors, acknowledging that authors are often working with software that itself wraps other software and therefore hides that software. Knepley et al~\cite{Knepley_WSSSPE} approach this by proposing a mechanism by which the software itself, after it has run, provides the user with a set of citations that are relevant to the code actually run. They describe a prototype implementation whereby the citations are embedded in libraries and reported along with the results, via a commandline interface~\cite{Knepley_WSSSPE}. Discussion highlighted the difficulty that attempting to acknowledge the contributions of all pieces of dependent code within a paper faces the difficulty of creating very long citation lists, straining the analogy of code used to papers cited. Katz approaches this issue by proposing a system of ``transitive credit,'' recording dependencies and relative contributions outside particular papers, relieving authors from the responsibility of acknowledging each and every dependency. Instead authors would acknowledge the percentage contribution of the software they used directly and an external system would then be able to recursively allocate that credit to those who had provided dependencies. Finally Priem and Piwowar argued that machine learning techniques could examine the body of published literature and extract mentions of software, coping with the multitude of informal ways in which authors mention software they have used~\cite{Priem_WSSSPE}.  Discussion included turning the emphasis from asking users to improve their citation of software contributions, to ask how projects producing software might monitor the literature to improve their ability to show impact. Michael McLennan described the approach taken by the NanoHub project to scan the literature using keywords and names of known users to discover papers that are likely to have used their software and platform, then to have graduate students read each paper, highlighting any mention and perhaps following up with the authors to locate stories of impact.  Work since the workshop has described this process at \textit{publications.wikia.com}.

Acknowledgement in science does not come only in publications, of course. A key location for visibility might be in the grant funding process, both in bio-sketches and in grant project reporting. Some progress has been made here. Representatives of the NSF at the meeting emphasized these opportunities and encouraged participants to take advantage of them. Nonetheless, there was skepticism that peer review panels would value these contributions in the same ways as publications.

Priem and Piwowar argued for additionally looking beyond publications and drawing on evidence of contribution, and impact, recorded in other online systems, such as source code repositories like Github, both code contributions, downloads of code and dependencies as well as conversations about software (mailing lists, twitter and beyond)~\cite{Priem_WSSSPE}. They argue for providing scholars with flexible resources so that they can tell their own stories in the manner most appropriate for them and their audiences, a principle of the ``altmetrics'' approach.
